---
title: "AMEC 2026 | Indiana University School of Medicine"
subtitle: "R for SNMA"
author: "Mirindi Kabangu"
format: 
  live-revealjs:
    self-contained: true
    footer: "IU School of Medicine | SNMA AMEC"
    logo: iu_logo.png # Optional: if you have the IU trident logo
    theme: [default]
    incremental: false
engine: knitr
webr:
  render-df: paged-table
  packages: [ 'readr', 'dplyr', 'tidyr', 'ggplot2', 'gt', 'gtsummary','ggsignif', 'palmerpenguins', 'medicaldata']
  cell:
    autoload: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ./_extensions/r-wasm/live/_gradethis.qmd >}}

## Instructor

Mirindi Kabangu

## Scope

::: aside
Taming the Data Beast, by Allison Horst
:::

## Overview

## The "Real-World" Medical Researcher Problem

You are a resident on a busy clinical rotation. You’ve been handed a
messy Vital Signs log from a 3-patient pilot study.

-   The Data:
-   The Goal:
    -   Download the CSV
    -   Split blood pressure into two columns
    -   Remove the "F" from Temperature
    -   Create a "Flag" column for any patient with a systolic BP \>
        140.5
    -   Make a bar chart of all the Systolic BPs
-   You have 2 minutes

## 

```{r}
#| label: reveal-cleaning
#| echo: false
library(flipbookr)
library(tidyverse)

# Secretly create the messy data for the demo
messy_vitals <- tibble::tibble(
  patient = c("Gemma Dunn", "Priscilla Holding", "Aniya Warner"),
  bp = c("128/78", "148/92", "115/73"),
  temp = c("98.6F", "101.2F", "97.5F")
)
```

##

```{r}
#| label: cleaning-demo-1
#| echo: true
#| output-location: column
#| code-line-numbers: "1"

print(messy_vitals)
```

##

```{r}
#| label: cleaning-demo-2
#| echo: true
#| output-location: column
#| code-line-numbers: "1-3"

messy_vitals |>
  # Separate BP into two numbers
  separate(bp, into = c("sys", "dia"), sep = "/")
```

##

```{r}
#| label: cleaning-demo-3
#| echo: true
#| output-location: column
#| code-line-numbers: "4-5"

messy_vitals |>
  # Separate BP into two numbers
  separate(bp, into = c("sys", "dia"), sep = "/") |>
  # Clean Temp units
  mutate(temp = parse_number(temp))
```

##

```{r}
#| label: cleaning-demo-4
#| echo: true
#| output-location: column
#| code-line-numbers: "6-7"

messy_vitals |>
  # Separate BP into two numbers
  separate(bp, into = c("sys", "dia"), sep = "/") |>
  # Clean Temp units
  mutate(temp = parse_number(temp)) |>
  # Convert types
  mutate(across(c(sys, dia), as.numeric))
```

##

```{r}
#| label: cleaning-demo-5
#| echo: true
#| output-location: column
#| code-line-numbers: "8-9"

messy_vitals |>
  # Separate BP into two numbers
  separate(bp, into = c("sys", "dia"), sep = "/") |>
  # Clean Temp units
  mutate(temp = parse_number(temp)) |>
  # Convert types
  mutate(across(c(sys, dia), as.numeric)) |>
  # Add the Clinical Flag
  mutate(flag = sys >= 140)
```

##

```{r}
#| label: cleaning-demo-7
#| echo: true
#| output-location: column
#| code-line-numbers: "10-12"

messy_vitals |>
  # Separate BP into two numbers
  separate(bp, into = c("sys", "dia"), sep = "/") |>
  # Clean Temp units
  mutate(temp = parse_number(temp)) |>
  # Convert types
  mutate(across(c(sys, dia), as.numeric)) |>
  # Add the Clinical Flag
  mutate(flag = sys >= 140) |>
  # Instant Visualization
  ggplot(aes(x = patient, y = sys, fill = flag)) +
  geom_col()
```

##

```{r}
#| label: cleaning-demo-8
#| echo: true
#| output-location: column
#| code-line-numbers: "13-14"

messy_vitals |>
  # Separate BP into two numbers
  separate(bp, into = c("sys", "dia"), sep = "/") |>
  # Clean Temp units
  mutate(temp = parse_number(temp)) |>
  # Convert types
  mutate(across(c(sys, dia), as.numeric)) |>
  # Add the Clinical Flag
  mutate(flag = sys >= 140) |>
  # Instant Visualization
  ggplot(aes(x = patient, y = sys, fill = flag)) +
  geom_col() +
  scale_fill_manual(values = c("gray", "#990000")) + # IU Crimson!
  theme_minimal()
```

## Another Challenge

You are a resident on a busy clinical rotation. You’ve been handed a
messy of cancer treatement data.

- The Data
- The Goal:
    - Download the CSV
    - Calculate the mean and standard deviations for each row
    - Calculate the p-values

## 

```{r}
#| label: table-demo-1
#| echo: true
#| output-location: column
#| code-line-numbers: "1"

library(gtsummary)

# Use the built-in 'trial' clinical dataset
gtsummary::trial |>
  select(trt, age, grade, response) |>
  # Create the summary table
  tbl_summary()
```

## 

```{r}
#| label: table-demo-2
#| echo: true
#| output-location: column
#| code-line-numbers: "1"

library(gtsummary)

# Use the built-in 'trial' clinical dataset
gtsummary::trial |>
  select(trt, age, grade, response) |>
  # Create the summary table
  tbl_summary(
    by = trt, # Split by treatment group
  )
```

## 

```{r}
#| label: table-demo-3
#| echo: true
#| output-location: column
#| code-line-numbers: "1"

library(gtsummary)

# Use the built-in 'trial' clinical dataset
gtsummary::trial |>
  select(trt, age, grade, response) |>
  # Create the summary table
  tbl_summary(
    by = trt, # Split by treatment group
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  )
```

## 

```{r}
#| label: table-demo-4
#| echo: true
#| output-location: column
#| code-line-numbers: "1"

library(gtsummary)

# Use the built-in 'trial' clinical dataset
gtsummary::trial |>
  select(trt, age, grade, response) |>
  # Create the summary table
  tbl_summary(
    by = trt, # Split by treatment group
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) |>
  # Add statistical comparison (p-values)
  add_p()
```

## 

```{r}
#| label: table-demo-5
#| echo: true
#| output-location: column
#| code-line-numbers: "1"

library(gtsummary)

# Use the built-in 'trial' clinical dataset
gtsummary::trial |>
  select(trt, age, grade, response) |>
  # Create the summary table
  tbl_summary(
    by = trt, # Split by treatment group
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) |>
  # Add statistical comparison (p-values)
  add_p() |>
  # Add a "Total" column
  add_overall()
```

## 

```{r}
#| label: table-demo-6
#| echo: true
#| output-location: column
#| code-line-numbers: "1"

library(gtsummary)

# Use the built-in 'trial' clinical dataset
gtsummary::trial |>
  select(trt, age, grade, response) |>
  # Create the summary table
  tbl_summary(
    by = trt, # Split by treatment group
    label = list(
      age ~ "Patient Age",
      grade ~ "Tumor Grade",
      response ~ "Treatment Response"
    ),
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) |>
  # Add statistical comparison (p-values)
  add_p() |>
  # Add a "Total" column
  add_overall() |>
  # Make it look professional
  bold_labels() |>
  modify_header(label = "**Variable**")
```

# Your Turn

:::{.fragment}
```{webr}
#| exercise: ex_2
#| setup: true
library(palmerpenguins)
library(medicaldata)
library(gtsummary)
penguins <- penguins[1:6,]
library(dplyr)
```

## Exercise 1: Isolate High-Grade Tumors {.smaller}

The `blood_storage` dataset contains a `Grade` variable (1-4). Filter the data to show only the most aggressive tumors (Grade 4) and assign it to `high_grade_cancers`.

:::{.fragment}
```{webr}
#| exercise: medex_1
# Filter for Grade 4 tumors
high_grade_cancers <- blood_storage |> 
  filter(Grade == ____)

# Count the results
nrow(high_grade_cancers)
```

::: { .solution exercise="medex_1"} 
::: { .callout-tip collapse=false}

```{webr}
#| exercise: medex_1
#| solution: true
high_grade_cancers <- blood_storage |> 
  filter(Grade == 4)

nrow(high_grade_cancers)
```

:::
:::

```{webr}
#| exercise: medex_1
#| check: true
gradethis::grade_this_code()
```

:::

---

### Exercise 2: Mutate and Clinical Thresholds (Intermediate)
**Goal:** Practice the "Rectangle Rule" by adding a new logical "Flag" column to indicate a clinical threshold (PSA level), as described in the Chen and Brown text.

## Exercise 2: Defining Clinical Risk {.smaller}

A PSA (Prostate Specific Antigen) level $\ge$ 10 is often considered high risk. Use `mutate()` to create a new column called `high_psa` that is `TRUE` if `PreopPSA` is 10 or greater.

:::{.fragment}

```{webr}
#| exercise: medex_2
# Create a logical flag for high PSA
psa_analysis <- blood_storage |> 
  mutate(high_psa = ________ >= 10)

# View the first few rows
head(psa_analysis)
```

::: { .solution exercise="med_ex_2"} 
::: { .callout-tip collapse=false}

```{webr}
#| exercise: medex_2
#| solution: true
psa_analysis <- blood_storage |> 
  mutate(high_psa = PreopPSA >= 10)

head(psa_analysis)
```

:::
:::

```{webr}
#| exercise: medex_2
#| check: true
gradethis::grade_this_code()
```

:::

---
### Exercise 3: The Multi-Step Pipeline (Advanced)
**Goal:** Combine multiple operations to prepare a dataset for a "Table 1" or summary report. This reinforces the idea that data cleaning is a series of explicit "recipe" steps.

## Exercise 3: Complex Data Preparation {.smaller}

The PI wants to look at younger patients (Age $\le$ 60) who had a cancer recurrence (`Recurrence == 1`). Select only those patients and calculate their average PSA.

:::{.fragment}
```{webr}
#| exercise: medex_3
# Combine filter and summarize
summary_stats <- blood_storage |> 
  filter(Age <= 60, ________ == 1) |> 
  summarize(mean_psa = ________(PreopPSA, na.rm = TRUE))

summary_stats
```

::: { .solution exercise="medex_3"} 
::: { .callout-tip collapse=false}

```{webr}
#| exercise: medex_3
#| solution: true
summary_stats <- blood_storage |> 
  filter(Age <= 60, Recurrence == 1) |> 
  summarize(mean_psa = mean(PreopPSA, na.rm = TRUE))

summary_stats
```

::: 
:::

```{webr}
#| exercise: medex_3
#| check: true
gradethis::grade_this_code()
```

:::